Stagnating State Funding for Higher Education and its Effect on Faculty Composition at US Universities

### 2023-05-31 text/sections/00-abstract.tex
Fixed spelling.

US universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while at the same time states have limited how much they fund higher education.
This analysis relates the changes the faculty composition at US public universities to the fall in state funding for higher education, using a shift-share approach to identify causal effects.
A fall of $-10$\% in university funding, via a state funding shock, decreases the number of assistant professors per student by $-1.4$\% and full professors by $-1.2$\%, yet increases the number of lecturers per student by 4.4\%; long-run estimates show the effects linger for multiple years after the initial funding shock.
Analysis of all the professors at Illinois public universities shows that incumbent professors are not affected by substantial funding shocks over 2011-2021, with suggestive evidence showing that the faculty changes arose by limiting the hiring of new professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

>>>>

US universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while at the same time states have limited how much they fund higher education.
This analysis relates the changes the faculty composition at US public universities to the fall in state funding for higher education, using a shift-share approach to identify causal effects.
A fall of $-10$\% in university funding, via a state funding shock, decreases the number of assistant professors per student by $-1.4$\% and full professors by $-1.2$\%, yet increases the number of lecturers per student by 4.4\%; long-run estimates show that the faculty composition is affected both in the same year as the funding shock, and two to three years following.
Analysis of all the professors at Illinois public universities shows that incumbent professors are not affected by substantial funding shocks over 2011-2021, with suggestive evidence showing that the faculty changes arose by limiting the hiring of new professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

### 2023-05-31 text/sections/04-empirical.tex
Fixed spelling.

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{Mean Characteristics for Public Universities, by State Funding Shock Instrument.}
    \makebox[\textwidth][c]{\input{tables/summary-quantiles.tex}}
    \label{tab:summary-quantiles}
    \footnotesize
    \textbf{Note}:
    The column labelled ``1st'' refers to the mean for all university-year observations in the first quintile (bottom 20\%) of the funding shock distribution, and so on.
    The mean of the funding shock within each quintile is shown in the first row of the table.
\end{table}

@@

The primary empirical model combines the instrumental variable for state funding shocks with the empirical model for the effects of state funding at the university-level --- i.e., parameter $\beta$ in the following.\footnote{
    It is important to note the treatment effect isolated here; the instrumental variables approach identifies the local average treatment effects, one specific to the instrument.
    So we interpret this treatment effect as a university's response in employment count and average salaries to state funding changes, changes specific to state funding shocks, among the complier group --- i.e., universities who respond to funding shocks who would not have made faculty-outcome changes absent the funding shock.
    Also, we assume that no universities' total revenues increase in response to negative state funding shock (monotonicity).
}
\begin{eqnarray}
    \label{eqn:secondstage1}
    X_{i,t} &=& \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t} \\
    \label{eqn:secondstage2}
    Y_{i,t} &=& \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{eqnarray}
I estimate the system by two stage least squares, including institution and year fixed effects, and investigate outcomes at the university-level to analyse effects on the university as a result of changes in revenues.
Additionally, I estimate the model via local projections \citep{jorda2005,miller2022} to investigate whether the effects on faculty composition linger for multiple years after the original funding shock.\footnote{
    The local projections method is an empirical method to estimate dynamic treatment effects when the treatment is not binary, so that time-series confounding (i.e., treatment in time $t$ is correlated with treatment in time $t-1$) is present, even for instrumental variable models.
    Inference for this model is similar to that for an event study design, so I present estimates in graphical form (see \autoref
    {fig:count-lp}).
}
Regarding outcomes, I focus on the composition of the professors employed at the university by analysing (log) count of professors per student by each university.

>>>>

\begin{table}[h!]
    \singlespacing
    %\centering
    \caption{Mean Characteristics for Public Universities, by State Funding Shock Instrument.}
    \makebox[\textwidth][c]{\input{tables/summary-quantiles.tex}}
    \label{tab:summary-quantiles}

    \footnotesize
    %\vspace{1mm}
    \textbf{Note}:
    The column labelled ``1st'' refers to the mean for all university-year observations in the first quintile (bottom 20\%) of the funding shock distribution, and so on.
    The mean of the funding shock within each quintile is shown in the first row of the table.
\end{table}

@@

The primary empirical model combines the instrumental variable for state funding shocks with the empirical model for the effects of state funding at the university-level --- i.e., parameter $\beta$ in the following.\footnote{
    It is important to note the treatment effect isolated here; the instrumental variables approach identifies the local average treatment effects, one specific to the instrument.
    So we interpret this treatment effect as a university's response in employment count and average salaries to state funding changes, changes specific to state funding shocks, among the complier group --- i.e., universities who respond to funding shocks who would not have made faculty-outcome changes absent the funding shock.
    Also, we assume that no universities' total revenues increase in response to negative state funding shock (monotonicity).
}
\begin{eqnarray}
    \label{eqn:secondstage1}
    X_{i,t} &=& \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t} \\
    \label{eqn:secondstage2}
    Y_{i,t} &=& \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{eqnarray}
I estimate the system by two stage least squares, including institution and year fixed effects, and investigate outcomes at the university-level to analyse effects on the university as a result of changes in revenues.
Additionally, I estimate the model via local projections \citep{jorda2005,miller2022} to investigate whether faculty composition is affected in the years after the initial funding shock (in addition to the initial year).\footnote{
    The local projections method is an empirical model used to estimate dynamic treatment effects when the treatment is not binary, so that time-series confounding (i.e., treatment in time $t$ is correlated with treatment in time $t-1$) is present, even for instrumental variable models.
    Inference in this method is for whether treatment $X_{i,t}$ affects the outcome $Y_{i,t+ k}$ in follow on years $k = 1, \hdots$, in addition to year $t$.
    I present estimates in graphical form (see \autoref{fig:count-lp}).
}
Regarding outcomes, I focus on the composition of the professors employed at the university by analysing (log) count of professors per student by each university.

### 2023-05-31 text/sections/05-results.tex
Fixed spelling.

Faculty composition is not only impacted in the same years as a state funding shock, but also for a number years following.
\autoref{fig:all-count-lp} shows local projection estimates, where the effect on total professor count per student persists at a magnitude of around 0.1 four years after the initial funding shock.
These findings are in line with the observation that universities froze hiring for a couple of yearsin response to 2008 negative budget shocks, particularly for tenure-track positions \citep{turner2014impact}, and similarly from shocks to university endowments \citep{brown2014endowment}.
For lecturers we see a negative effect, which lines up with two trends noted in \autoref{sec:trends}: public universities revenues (per student) are decreasing at the same time as employment of contingent lecturers increased rapidly.
And yet, identifying the university revenues by the state funding shock increases the magnitude of the effect (Column 2) with respect to the association (Column 1), strengthening the case that the two trends are causally linked.\footnote{
    Between years 1990-1999 IPEDS provided count of professors by whether they had tenure (opposed to named position as in \autoref{tab:facultycount-shock-reg}), and \autoref{tab:tenurecount-shock-reg-fte} presents results of the respective regressions by tenure status.
    Results are relatively similar, yet imprecise from the much smaller number of observations.
}

>>>>

Faculty composition is not only impacted in the same years as a state funding shock, but also for a number years following.
\autoref{fig:all-count-lp} shows local projection estimates, where the effect on total professor count per student is around 0.1 in both the same year year as the initial funding shock, and for the next 2 years
These findings are in line with the observation that universities froze hiring for a couple of years in response to 2008 negative budget shocks, particularly for tenure-track positions \citep{turner2014impact}, and similarly from shocks to university endowments \citep{brown2014endowment}.
For lecturers, we see a negative effect which lines up with two trends noted in \autoref{sec:trends}: public universities revenues (per student) are decreasing at the same time as employment of contingent lecturers increased rapidly.
And yet, identifying the university revenues by the state funding shock increases the magnitude of the effect (Column 2) with respect to the association (Column 1), strengthening the case that the two trends are causally linked.\footnote{
    Between years 1990-1999 IPEDS provided count of professors by whether they had tenure (opposed to named position as in \autoref{tab:facultycount-shock-reg}), and \autoref{tab:tenurecount-shock-reg-fte} presents results of the respective regressions by tenure status.
    Results are relatively similar, yet imprecise from the much smaller number of observations.
}
